Red Wine Analysis by Gordon Yu
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
 red_wine <- read.csv('wineQualityReds.csv')

```

This dataset contains information about 1599 different kinds of red wine. There
are 11 different attributes, along with one output variable called quality. 

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
dim(red_wine)
```

```{r echo=FALSE}
str(red_wine)
```

```{r echo=FALSE}
summary(red_wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=factor(quality)))+
  geom_bar()
```

First, I looked at the distribution of the quality variable. The text file 
describing the dataset said that the quality variable ranges from 0 to 10. 
Based on this graph, it looks like most of the wines have a quality of 5 or 6. 
I'm surprised by how low the average quality rating is and how the highest 
quality value in this dataset is 8. I thought there would be at least 1 type 
of red wine that would score a 10 for quality. 


```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=fixed.acidity))+
  geom_histogram()
```

Fixed acidity is skewed to the right. Most wines have a fixed acidity between 
6 and 8.

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=fixed.acidity))+
  geom_boxplot()
```

After plotting a boxplot for fixed acidity, We can now see that the wines 
with fixed acidity of around 13 or higher are outliers. The median of fixed
acidity is around 8 and the third quartile is around 9. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=volatile.acidity))+
  geom_histogram()
```

Volatile acidity is also skewed to the right. It seems like the distribution for
volatile acidity is pretty similar to the distribution of fixed acidity. It
makes sense since they're both measures of acidity. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=volatile.acidity))+
  geom_boxplot()
```

We can see that the outliers for volatile acidity span from the range of 
 1.0 to a little below 1.6. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=citric.acid))+
  geom_histogram()
```

The histogram for the citric acid is skewed to the right. The most common amount
of citric acid in red wines is around 0, so it seems like most red wines contain
very small amounts of citric acid. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=citric.acid))+
  geom_boxplot()
```

It looks like there's actually only one outlier for citric acid and it's at the
value of 1.0. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=residual.sugar))+
  geom_histogram()
```

The histogram for residual sugar is very skewed to the right, with most red 
wines having less than 4 g/dm^3 of sugar. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=residual.sugar))+
  geom_boxplot()
```

The boxplot shows us that there are lot of outliers for residual sugar.
The range of the outliers is very wide, starting from around 4 and going up
to almost 16. Most of the outliers are between the values of 4 and 9. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=chlorides))+
  geom_histogram()
```

The graph for chlorides is also really skewed to the right. It looks like the 
distributions for residual sugar and chlorides are really similar.I think that
is interesting since sugar and salt(chlorides) are opposites. Most red wines 
have less than 0.1 sodium chloride - g / dm^3.

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=chlorides))+
  geom_boxplot()
```

This boxplot for chlorides is very interesting. The IQR is extremely small. 
There's also a large number of outliers, with the majority of them larger than
the third quartile value and a few are lower than the first quartile value. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=free.sulfur.dioxide))+
  geom_histogram()
```

The free sulfur dioxide is also skewed to the right. The most common free sulfur
dioxide amount is around 5 mg/dm^3

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=free.sulfur.dioxide))+
  geom_boxplot()
```

The outliers for free sulfur dioxide range from 45 to 75. The median is around 
15. The first and third quartiles look like they're almost the same distance
away from the median. 

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=total.sulfur.dioxide))+
  geom_histogram()
```

Total sulfur dioxide is also skewed to the right.

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=total.sulfur.dioxide))+
  geom_boxplot()
```

The outliers for total sulfur dioxide range from 125 to almost 300, which is a
pretty big range. Most of the outliers are between the values of 125 and 175, 
with a lot of overplotting in that area. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=density))+
  geom_histogram()
```

The distribution of density looks pretty close to normal. It also looks like the
range of values is relatively small. Most wines have a density between 0.995
and 1.0 g / cm^3. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=pH))+
  geom_histogram()
```

The distribution of pH also looks pretty close to normal. It also seems like all
of the wines are relatively acidic since they're mainly between the values of 3
and 4 on the pH scale. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=sulphates))+
  geom_histogram()
```

The distribution for sulphates is skewed to the right. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=sulphates))+
  geom_boxplot()
```

The outliers for sulphates range from 1.0 to 2.0. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=alcohol))+
  geom_histogram()
```

The distribution of alcohol is skewed to the right. It looks like most of the
wines have an alcohol percentage between 9% and 10%. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(y=alcohol))+
  geom_boxplot()
```

There are only a few outliers for alcohol that range from 13.5 to 15. 

Now I'll create some new variables to separate the residual sugar and pH
variables into different levels to help with further analysis. 

```{r echo=FALSE}
#get the different quantile values for residual sugar and pH
summary(red_wine$residual.sugar)
summary(red_wine$pH)

```

```{r echo=FALSE}
#create the different bin labels
sugar_bins <- c('low', 'moderate', 'medium', 'high')

acidity_bins <- c('high', 'medium', 'moderate', 'low')

#create different bins for residual sugar variable
red_wine$sugar_levels <- cut(red_wine$residual.sugar, 
                        breaks=c(0.9, 1.9, 2.2, 2.6, 15.5), labels= sugar_bins)

#create different bins for pH variable
red_wine$acidity_levels <- cut(red_wine$pH,breaks=c(2.74, 3.21, 3.31, 3.4, 4.01)
                               , labels= acidity_bins)


```

After creating the new variables, I will now plot the histograms for them. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=sugar_levels))+
  geom_bar()
```

We can see that the wines with the lowest sugar are the most common. There's 
almost 100 more low sugar wines compared to high sugar wines. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=acidity_levels))+
  geom_bar()
```

The high acidity level is the most common but there isn't a big difference 
between the counts for the different levels. 

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 different types of red wine in the datase and 12 different 
variables that track the different attributes. 

### What is/are the main feature(s) of interest in your dataset?

The main feature in this dataset is quality.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I want to see if the values of residual sugar, pH, and alcohol affect the
quality of the wines. 

### Did you create any new variables from existing variables in the dataset?

I created some variables that I will use in the Multivariate plots section to
separate the residual sugar and pH values into different levels for a better 
analysis. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The most unusual thing to me was the distribution of the quality ratings for the
wines. I didn't expect the majority of wines to have quality ratings of 5 or 6. 
I thought there would be some wines that have a rating of 9 or 10 but the 
highest quality rating is actually 8. 


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots}
ggplot(red_wine, aes(x=quality, y=alcohol))+
  geom_point()
```

First, I wanted to look at the relationship between quality and alcohol 
percentages. 

It looks like wines with a quality rating of 5 have the largest range of alcohol
percentage values, with the highest alcohol percentage around 15%. There's also
some overplotting. 

```{r echo=FALSE}
ggplot(red_wine, aes(x=quality, y=alcohol))+
  geom_point(alpha=0.2, position='jitter')
```

After changing the alpha value for the plotted points, a few noticeable things
pop up. It looks like there's a large cluster of wines that have a quality 
rating around 5 and contain a 9% alcohol. On the other hand,wines with the 
second most common quality rating of 6 look like they have alcohol percentages
that are a little bit more spread out. 

```{r echo=FALSE}
ggplot(red_wine, aes(x=quality, y= residual.sugar))+
  geom_point()
```

Similar to the last scatterplot, we see that the wines with the most common 
quality ratings of 5 and 6 have the largest outliers. There's also some 
overplotting.

```{r echo=FALSE}
ggplot(red_wine, aes(x=quality, y= residual.sugar))+
  geom_point(alpha=0.2, position='jitter')
```

After changing the alpha value for the points and adding jitter, we can see what
there's two large clusters for the quality ratings of 5 and 6 around the 
residual sugar amount of 2 g / dm^3. It leads me to believe that most wines
contain a very small amount of residual sugar. 




When we look at the average amount of residual sugar for each quality rating, we
see that wines with a quality rating of 6 have the lowest average. I don't think
that the graph reveals much though. 

```{r echo=FALSE}
ggplot(red_wine, aes(x=quality, y= pH))+
  geom_point()
```

The most acidic wine in the dataset has a quality rating of 4. The graph is also
very overplotted. 

```{r echo=FALSE}
ggplot(red_wine, aes(x=quality, y= pH))+
  geom_point(alpha=0.2, position = 'jitter')
```

After changing the alpha value for the points and adding jitter, we can see that
there's a large cluster of wines witg quality ratings of 5 and 6 that have
pH values around 3.25 - 3.5. 

```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=residual.sugar, y= chlorides))+
  geom_point()

```

It looks like most wines contain low amounts of chlorides and residual sugar. 
It makes sense given the histograms of chlorides and residual sugar from before
that show large counts for the smallest amounts. There are also some outliers
that contain relatively high amounts of chlorides and low amounts of residual
sugar and some that have the opposite relationship. 

```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=residual.sugar, y= chlorides))+
  geom_point(alpha=0.1, position='jitter')+
  xlim(0, quantile(red_wine$residual.sugar, 0.95))+
  ylim(0, quantile(red_wine$chlorides, 0.95))

```


After getting rid of some of the outliers and changing the alpha value for the
points, we can see a cluster of points around the 0.075 value for chlorides. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=pH, y= density))+
  geom_point()
```

```{r echo=FALSE}
cor(red_wine$pH, red_wine$density)
```

Based on the scatterplot and the correlation coefficient, it seems like pH
and density have a moderate negative relationship. It also looks like there's a
large cluster of points between the pH values of 3.25 and 3.5 and the density 
values of 0.995 and 0.9975. 

```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=pH, y= density))+
  geom_point(alpha=0.2, position='jitter')+
  xlim(2.7, quantile(red_wine$pH, 0.95))+
  ylim(0.99, quantile(red_wine$density, 0.95))

```

When we zoom in to the graph and change the alpha value of the points, it 
doesn't reveal much. We just get a slightly better look at the clusters of
points. 

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Most of the variables don't seem to have much of a linear relationship with
each other. 

There is a moderately strong linear relationship between density and pH. 
There are a lot of wines that have density and pH values that are in the middle
of the ranges for those variables. 

There are a lot of wines with quality ratings of 5 and 6 that have low alcohol
percentages and low residual sugar. There also seems to be a lot of wines that
have low residual sugar and low chlorides. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There wasn't any other interesting relationships. 

### What was the strongest relationship you found?

The strongest relationship seemed to be from density vs. pH. The graph had the
most linear relationship and the correlation coefficient confirmed that there 
is a moderate linear relationship between the two variables. 

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}
ggplot(red_wine, aes(x= quality, y= alcohol))+
  geom_point(aes(color=red_wine$sugar_levels))
```

After adding the different sugar levels as a third variable to the graph of
alcohol vs. quality, we can see that the wine with the highest alcohol 
percentage also has a high sugar level. The wine with the combination of the 
lowest alcohol percentage and lowest quality has a moderate sugar level. 
Howeve, there are also two other wines with similar alcohol percentages but 
higher quality ratings that have low sugar levels. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= quality, y= alcohol))+
  geom_point(alpha=0.2, position = 'jitter' ,aes(color=red_wine$sugar_levels))
```

When we change the alpha value and add some jitter, it doesn't reveal any new
insights. 

```{r echo=FALSE,  message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= quality, y= alcohol) )+
  geom_point(aes(color=red_wine$acidity_levels))
```

After adding the different acidity levels as a third variable to the graph of
alcohol vs. quality, we can see that the wines with the lowest and highest
alcohol percentages all have high acidity levels. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= quality, y= alcohol) )+
  geom_point(alpha= 0.2, position= 'jitter' , aes(color=red_wine$acidity_levels))
```

When we change the alpha value and add some jitter, it doesn't reveal any new
insights. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= quality , y= pH))+
  geom_point(aes(color=red_wine$sugar_levels))
```

After adding the different sugar levels as a third variable to the graph of
pH vs. quality, there doesn't seems to be any noticeable patterns. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= quality , y= pH))+
  geom_point(alpha = 0.2, position = 'jitter', aes(color=red_wine$sugar_levels))
```

When we change the alpha value and add some jitter, it doesn't reveal any new
insights. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x= residual.sugar , y= chlorides))+
  geom_point( aes(color=red_wine$acidity_levels))
```

Adding the acidity level to the graph of chlorides vs. residual sugar doesn't 
reveal much. A lot of the most extreme outliers in this graph have high acidity 
levels. 

```{r echo=FALSE ,message=FALSE, warning=FALSE}
ggplot(red_wine, aes(x=residual.sugar, y= chlorides))+
  geom_point(alpha=0.15, position='jitter', aes(color=red_wine$acidity_levels))+
  xlim(0, quantile(red_wine$residual.sugar, 0.95))+
  ylim(0, quantile(red_wine$chlorides, 0.95))

```

After changing the alpha value and getting rid of some of the outliers, we
still can't really see any clear patterns. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I found that the wines with the lowest and highest alcohol percentages all have
high acidity levels. I also found out that many of the wines that have high 
amounts of chlorides and low amounts of sugar also have high acidity levels. 

### Were there any interesting or surprising interactions between features?

There wasn't anything that really surprised me. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(red_wine, aes(x=factor(quality)))+
  geom_bar()+
  xlab('quality')+
  ylab('Number of Wines')+
  ggtitle('Red Wine Quality')
  #theme_update(plot.title = element_text(hjust = 0.5))
```


### Description One

This graph shows you that most of the wines have a quality rating of 5 or 6.
I chose this graph because I thought the distribution was very interesting. 
I expected there to be more wines with higher ratings. I also though that the
highest quality rating would be a 9 or a 10. 

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(red_wine, aes(y=chlorides))+
  geom_boxplot()+
  ggtitle('Chlorides Boxplot')+
  theme(axis.text.x=element_blank())
```


### Description Two

This boxplot was the most interesting one out of all the ones that I plotted. 
It shows you that the The IQR is extremely small. 
There's also a large number of outliers, with the majority of them larger than
the third quartile value and a few are lower than the first quartile value.  

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(red_wine, aes(x=pH))+
  geom_histogram()+
  ggtitle('Red Wine pH')+
  ylab('Number of wines')

```

### Description Three

This third graph shows you the counts of the different pH values for the wines.
I picked this graph because pH is one of the few variables that has a close
to normal distribution and I thought that was a little interesting. You can see
that the most common pH value is around 3.3. 

------

# Reflection

This dataset contained info about 1599 different kinds of red wine. I first
looked at each variable individually by plotting histograms. I saw that the
most common quality rating was 5 and 6. I also saw that a lot of the variables
were very skewed and had a lot of outliers. 

When I looked at the bivariate relationships, I saw that there is a moderately
strong linear relationship between density and pH. The other pairs of variables
didn't seem to have any real relatioships between each other. 

When I looked at the multivariate relatioships, I couldn't really uncover 
anything useful. 

I had a hard time finding linear relationships between different variables. I 
feel like adding an extra variable to a bivariate graph didn't really help and 
it was just hard to find any patterns. 

I think I did an ok job dealing with overplotting in the bivariate graphs. 

To expand on my analysis of this dataset, I would merge this red wine dataset
with the white wine dataset and analyze the similarities and differences between
the two. 

